Solar-Electric and Gas Powered, Long-Endurance UAV Sizing via Geometric Programming

### 2017-01-24 gassolarpaper/gassolarpaper.tex
restructure with requirements section

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{mtowvsendurance.pdf}
    \caption{ \textbf{ Trade study showing how gas powered aircraft weight increases with endurance.  }}
	\label{f:mtowvsendurance}
	\end{center}
\end{figure}

To capture the key trade offs like those presented in Figures~\ref{f:solarstart} and ~\ref{f:mtowvsendurance} a geometric programm is built that models for both gas and solar-electric powered, long-endurance aircraft.  
Equations governing the underlying physics of aerodynamic performance, solar energy collection and storage, and fuel burn are written in a geometric programming form, which can then be solved as an optimization problem.
Then by changing the design parameters, such as endurance, latitude, and percentile wind speeds, multiple solutions can be rapidly generated to observe how the aircraft size and weight is affected by different requirements.

@@

@@

\begin{align} 
\label{e:gpform}
\text{minimize } f_0(\bold{x}) & \nonumber \\
\text{subject to  } f_i(\bold{x}) &\leq 1, i=1,...,m \\
g_i (\bold{x}) &= 1, i = 1,...,p \nonumber 
\end{align}

@@

\subsection{Station Keeping}

The design problem considered is a long-endurance, station keeping aircraft or an aircraft capable of remaining in one location in the air for long periods of time. 
Long-endurance, station keeping aircraft are generally required to operate at specific latitudes, altitudes, and percentile wind speeds.\cite{zephyr} 
Both wind speeds and atmospheric quantities depend on the station keeping requirements.

\subsubsection{Wind Speeds}

To station keep, an aircraft must fly faster than the wind speed.  
Therefore, the aircraft speed is constrained for both the gas and solar-electric powered cases,

An analysis was done of how wind speeds vary from month to month to determine during which seasons or months the wind speeds are higher. Figure~\ref{f:windvsmonth} reveals that wind speeds are significantly worse during the winter months. 
Because the constraining case for solar-electric powered aircraft is the winter solstice and because the wind speeds are highest during the winter months only the wind speed data from December will be used for aircraft the sizing and performance analysis presented in this paper.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{windvsmonth.pdf}
    \caption{ \textbf{ Variation of wind speed by month at 45 degrees latitude at altitudes of 15,000 ft, 50,000 ft and 60,000 ft.  Left bound, right bound and middle lines represent 85th, 95th, and 90th percentile winds respectively. }}
	\label{f:windvsmonth}
	\end{center}
\end{figure}

Figure~\ref{f:latvswind} shows how the wind speed varies with latitude in the Northern Hemisphere. 
Because the gas-powered aircraft is assumed to have a naturally aspirated engine, it will fly between 5,000 and 25,000 ft depending on payload requirement. 
The solar-electric powered aircraft will operate around 60,000 ft to avoid cloud coverage.  
Variations of wind speed with latitude are shown for altitudes at both 15,000 ft at and 60,000 ft.  \\

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{latvswind.pdf}
    \caption{ \textbf{ Variation of wind speed with latitude at altitudes of 15,000 ft, 50,000 ft and 60,000 ft.  Left bound, right bound and middle lines represent 85th, 95th, and 90th percentile winds respectively. Wind data includes winds during December for the year 2005-2015.}}
	\label{f:latvswind}
	\end{center}
\end{figure}

Wind speeds also vary significantly with altitude. (Figure~\ref{f:altvswind})
The high winds around 30,000-40,000 ft makes it difficult for both solar-electric and gas powered long-endurance aircraft to fly in that range of altitudes.  
Solar-electric powered aircraft, that do not have aspirated engines, fly around 60,000 ft take advantage of lower wind speeds and limited cloud coverage.
Gas powered aircraft with aspirated engines can have difficulty operating at 60,000 ft and generally perform better in the lower wind speeds below 25,000 ft.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{altvswind.pdf}
    \caption{ \textbf{ Variation of wind speed with altitude at latitudes of 30, 35, and 45 degrees.  Left bound, right bound and middle lines represent 85th, 95th, and 90th percentile winds respectively. Wind data includes winds during December for the year 2005-2015.}}
	\label{f:altvswind}
	\end{center}
\end{figure}

To understand what a certain percentile wind speed implies operationally an autocorrelation analysis was done for multiple cites at varying latitudes at 60,000 during December. 
Boston and Rome during December of 2015 are shown in Figures~\ref{f:bostonwinds} and~\ref{f:romewinds}. 
One lag is equivalent to 6 hours.

\subsubsection{Optimizing Altitude for Solar-Electric Powered Aircraft}

@@

Because wind speeds also vary with latitude several constraints are needed to optimize altitude. 
Operationally, aircraft, such as the solar-electric powered aircraft, will fly within a band of latitudes (i.e. $\pm$45 degrees latitude).  
For example, an aircraft optimized for the 35th latitude would be able to operate at any latitude between $\pm35$ degrees latitude. 
This is imposed in the optimization model as multiple constraints. For the 35 degree latitude case, 70 separate wind and solar flux constraints that must be satisfied, one for each latitude.

Constraints relating wind speed and air density were generated using the data fitting technique described by Hoburg\cite{fitting}.
Air density was chosen as the dependent variable instead of altitude to eliminate an extra equation relating altitude to air density. 
Altitude can be post-calculated from the optimized air density. 
Because wind speeds are low and solar flux is higher below 20 degrees latitude no constraints for latitudes below 20 degrees are actually included in the optimization. 
Additionally, only wind speed data from the Northern Hemisphere are considered.
Therefore, 41 wind speed constraints were generated from 20-60 degrees North latitude. (See Appendix A) 
The constraint of wind speed and air density at the 35 degree latitude is

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{windfitl35.pdf}
    \caption{ \textbf{ Comparison of a fitted equation of wind speed to air density and percentile wind speed~\eqref{e:windfit35}. (Log space RMS error = 0.043.) Wind speed data includes December wind speeds from 2005-2015 at the 35th degree latitude. }}
	\label{f:windfitl35}
	\end{center}
\end{figure}

@@

\subsubsection{Atmospheric Quantities}

For the solar aircraft air density is being optimized and the operating altitude is computed after the fact using equations specific to the tropopause,\cite{isaatm}

Because altitude is a requirement for the gas powered aircraft, density is pre-computed prior to solving the optimization using atmospheric equations specific to the troposphere,\cite{isaatm}

@@

The driving equations for the solar-electric powered aircraft are determined by the energy required to fly throughout the day while charging the batteries enough to last through the night.  
Equations~\ref{e:solarreq} and~\ref{e:solarbatt} capture the solar requirement by enforcing that the total solar energy harvested by the solar cells during the day must be greater than the energy used to operate the aircraft during the day while charging enough batteries to fly during the night.\cite{solartech}
Under these assumptions, if the aircraft can fly for one day and one night, then it could theoretically fly for many days under similar circumstances.

    \begin{align}
        \label{e:solarreq}
        (E/S)_{\text{sun}} \eta_{\text{solar}} S_{\text{solar}} &\geq P_{\text{oper}}t_{\text{day}} + E_{\text{battery}}/\eta_{\text{charge}} \\
        \label{e:solarbatt}
        E_{\text{battery}} &\geq P_{\text{oper}}t_{\text{night}}/\eta_{\text{discharge}}
    \end{align}

    It is assumed that the charge and discharge efficiencies are constant, $\eta_{\text{charge}} = \eta_{\text{discharge}} = 0.98$. 
    It is also assumed that the solar cell efficiency is constant, $\eta_{\text{solar}} = 0.3$. 
    A battery with an energy density of $h_{\text{batt}} = 350$ [Whr/kg], is selected for the solar-powered aircraft.

    To enforce these constraints, two variables were created: $P_{\text{oper}}$ and $S_{\text{solar}}$. 
    $P_{\text{oper}}$ is defined as the average power required to operate the aircraft. 
    $S_{\text{solar}}$ is defined as the amount of area on the aircraft covered by solar cells.

    \begin{align}
        \label{e:solarpoper}
        P_{\text{oper}} &\geq P_{\text{shaft}} + P_{\text{avionics}} \\
        \label{e:solarssolar}
        S_{\text{solar}} &\leq f_{\text{solar}}S
    \end{align}

    For the purposes of this design study, it is assumed that the solar cells are only placed on the wing and that the total solar cell area is limited to some fraction of the wing planform area $f_{\text{solar}}$, to account for control surfaces, leading edge, and other accessories.

    Because solar aircraft requirements are typically given in terms of latitude and yearly availability it is useful to express the total available solar energy per unit area per day as a function of latitude, $\phi$, and the day of the year, $DOY$,

    \begin{equation}
        \label{e:solarfunc}
        (E/S)_{\text{sun}} = f(\phi, DOY).
    \end{equation}

    For simplicity, energy and power will refer to energy and power per unit area throughout the rest of this section (i.e. $(E/S) = E$). 
    The total available solar energy per day is an integral of the available solar power during the course of the day,

    \begin{equation}
        \label{e:solares}
        E_{\text{sun}} = \int_{t_{\text{sun rise}}}^{t_{\text{sun set}}} P_{\text{sun}} dt.
    \end{equation}

    Figure~\ref{f:lat45} shows a sample distribution of the available solar power during the spring equinox at 45 degrees latitude.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{lat30.pdf}
    \caption{ \textbf{ The available solar power per unit area during the winter solstice in the Northern Hemisphere at $\phi=30^{\circ}$.  The area under this curve is the total energy available per unit area for the entire day.} }
	\label{f:lat45}
	\end{center}
\end{figure}

The shape of this curve is governed by the angle, $\theta$, between the normal to the flat surface, or aircraft wing, and the sun beam.\cite{solar}

>>>>

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{mtowvsendurance.pdf}
    \caption{\textbf{Trade study showing how gas powered aircraft weight increases with endurance.}}
	\label{f:mtowvsendurance}
	\end{center}
\end{figure}

To capture the key trade offs like those presented in Figures~\ref{f:solarstart} and~\ref{f:mtowvsendurance} a geometric programm is built that models for both gas and solar-electric powered, long-endurance aircraft.  
Equations governing the underlying physics of aerodynamic performance, solar energy collection and storage, and fuel burn are written in a geometric programming form, which can then be solved as an optimization problem.
Then by changing the design parameters, such as endurance, latitude, and percentile wind speeds, multiple solutions can be rapidly generated to observe how the aircraft size and weight is affected by different requirements.

@@

\section{Requirements Analysis}

Design requirements to operate at specific altitudes, latitudes, and percentile wind speeds each of which will affect the performance and design of the aircraft.\cite{zephyr} 
To station keep an aircraft must fly faster than the wind speed,

\subsection{Endurance}

Long-endurance gas powered aircraft, the endurance requirement the amount of time the aircraft needs to remain in the air without refueling.  
If solar-electric powered aircraft can fly during the shortest day of the year, December 21st, then it can theortically fly at any time the year.  
Therefore, the endurance requirement for solar-electric powere aircaft is assumed to be a requirement to fly during the winter solstice.  
To achieve this design constraint the aircraft must be able to fly with limited solar energy, discussed in Section , and with higher wind speeds as shown in Figure~\ref{f:windvsmonth}.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{windvsmonth.pdf}
    \caption{\textbf{Variation of wind speed by month at 45 degrees latitude at altitudes of 15,000 ft, 50,000 ft and 60,000 ft.  Lower bound, upper bound and middle lines represent 85th, 95th, and 90th percentile winds respectively.}}
	\label{f:windvsmonth}
	\end{center}
\end{figure}

\subsection{Altitude}

Gas powered aircraft generally fly at a given altitude to meet a payload requirement.  
A naturally aspirated engine and linearly increasing wind speeds from 0 to 35,000 ft shown in Figure~\ref{f:altvswind} mean that gas powered aircraft will generally be more efficiect at lower altitude and generally fly between 5,000 at 25,000 ft depending on the payload requirement.  
For this design study, it is assume d that the aircraft is flying at 15,000 ft.

Solar-electric powered aircraft tend to fly around 60,000 ft to meet payload requirements, avoid cloud coverage and fly in lower wind speeds which have a minimum around 60,000 shown in Figure~\ref{f:altvswind}

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{altvswind.pdf}
    \caption{\textbf{Variation of wind speed with altitude at latitudes of 30, 35, and 45 degrees.  Left bound, right bound and middle lines represent 85th, 95th, and 90th percentile winds respectively. Wind data includes winds during December for the year 2005-2015.}}
	\label{f:altvswind}
	\end{center}
\end{figure}

The altitude also determines atmospheric properties such as air density and viscosity.  For the solar aircraft air density is calculated using equations specific to the tropopause,\cite{isaatm}

For the gas powered aircraft air density is calculated using atmospheric equations specific to the troposphere,\cite{isaatm}

\subsection{Latitude}

Latitude affects both the solar-electric and the gas powered aircraft because wind speed varies with latitude. 
Figure~\ref{f:latvswind} shows how the wind speed varies with latitude in the Northern Hemisphere.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{latvswind.pdf}
    \caption{\textbf{Variation of wind speed with latitude at altitudes of 15,000 ft, 50,000 ft and 60,000 ft.  Left bound, right bound and middle lines represent 85th, 95th, and 90th percentile winds respectively. Wind data includes winds during December for the year 2005-2015.}}
	\label{f:latvswind}
	\end{center}
\end{figure}

Latitude additionally affects the solar-electric powered aircraft because at higher latitudes there is less daylight and therefore less solar energy during the winter months. This is further discussed in Section.

Operationally, long-endurance aircraft will fly within a band of latitudes (i.e. $\pm$45 degrees latitude).  
For example, an aircraft optimized for the 35th latitude would be able to operate at any latitude between $\pm35$ degrees latitude. 
This can be imposed in the optimization model as multiple constraints. For the 35 degree latitude case, 70 separate wind and solar flux constraints that must be satisfied, one for each latitude.

\subsection{Percentile Wind Speed}

For the purposes of this design study, a percentile wind speed corresponds to that percentile of wind speed data in December across all longitudes and spanning the years 2005-2015.  
To understand what a certain percentile wind speed implies operationally an autocorrelation analysis was done for multiple cites at varying latitudes at 60,000 during December. 
Boston and Rome during December of 2015 are shown in Figures~\ref{f:bostonwinds} and~\ref{f:romewinds}.

@@

\begin{align} 
\label{e:gpform}
\text{minimize } f_0(\bold{x}) & \nonumber \\
\text{subject to  } f_i(\bold{x}) &\leq 1, i=1,\ldots,m \\
g_i (\bold{x}) &= 1, i = 1,\ldots,p \nonumber 
\end{align}

@@

\subsection{Wind Speeds}

@@

Because wind speeds also vary with latitude several constraints are needed to optimize altitude. 
Constraints relating wind speed and air density were generated using the data fitting technique described by Hoburg\cite{fitting}.
Air density was chosen as the dependent variable instead of altitude to eliminate an extra equation relating altitude to air density. 
Altitude can be post-calculated from the optimized air density. 
Because wind speeds are low and solar flux is higher below 20 degrees latitude no constraints for latitudes below 20 degrees are actually included in the optimization. 
Additionally, only wind speed data from the Northern Hemisphere are considered.
Therefore, 41 wind speed constraints were generated from 20-60 degrees North latitude. (See Appendix A) 
The constraint of wind speed and air density at the 35 degree latitude is

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{windfitl35.pdf}
    \caption{\textbf{Comparison of a fitted equation of wind speed to air density and percentile wind speed~\eqref{e:windfit35}. (Log space RMS error = 0.043.) Wind speed data includes December wind speeds from 2005-2015 at the 35th degree latitude.}}
	\label{f:windfitl35}
	\end{center}
\end{figure}

@@

@@

At a given time of year DOY, and latitude $\phi$, there exists an availble solar irradiated energy $(E/S)_{\text{sun}}$ per unit area during one day

\begin{equation}
    \label{e:solarfunc}
    (E/S)_{\text{sun}} = f(\phi, \text{DOY}).
\end{equation}

To operate continually for multiple days the total available solar energy $(E/S)_{\text{sun}}$, must be greater than the required solar energy per unit area to power the aircraft during the day $(E/S)_{\text{day}}$ and the energy required to power the aircraft during the night via batteries $E_{\text{batt}}$,\cite{solartech}

\begin{align}
    \label{e:solarreq}
    (E/S)_{\text{sun}}  &\geq (E/S)_{\text{day}} + \frac{E_{\text{battery}}}{\eta_{\text{charge}}\eta_{\text{solar}} S_{\text{solar}}} \\
    \label{e:solarbatt}
    E_{\text{battery}} &\geq \frac{P_{\text{oper}}t_{\text{night}}}{\eta_{\text{discharge}}} + (E/S)_C \eta_{\text{solar}} S_{\text{solar}}
\end{align}

where the charge, discharge and solar cell efficiencies are assumed to be $\eta_{\text{charge}} = \eta_{\text{discharge}} = 0.98$ and $\eta_{\text{solar}}= 0.22$. 
An additional energy term $(E/S)_C$ is included in Equation~\ref{e:solarbatt} to account for hours of low solar irradiance, during mornings and evenings, when the aircraft must continue to fly on battery power. 
Equations~\ref{e:solarreq} and~\ref{e:solarbatt} are graphically represented in Figure~\ref{f:lat30}, which represents the solar energy at 30 degrees latitude.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{lat30.pdf}
    \caption{\textbf{The available solar power per unit area during the winter solstice in the Northern Hemisphere at $\phi=30^{\circ}$. For a solar-electirc powered aircraft to be operational for more than 24 hours, the area under the red curve must be greater than the areas shown in blue, magenta and green.}}
	\label{f:lat30}
	\end{center}
\end{figure}

In order for the aircraft to begin charging batteries there must be a minimum solar irradiance power

\begin{align}
    (P/S)_{\text{min}} &= \frac{P_{\text{oper}}}{\eta_{\text{solar}} S_{\text{solar}}} \\
    \eta_{\text{motor}} P_{\text{oper}} &\geq P_{\text{shaft}} + P_{\text{avionics}}. \\
\end{align}

where $\eta_{\text{motor}} = 0.95$. Thus for a given latitude and day, and therfore total solar energy $(E/S)_{\text{sun}}$, both $(E/S)_{\text{day}}$ and $(E/S)_C$ are functions of $(P/S)_{\text{min}}$.  
The total solar energy available $(E/S)_{\text{sun}}$, is computed prior to the optimization solve and is used to generate monomial approximations for $(E/S)_{\text{day}}$ and $(E/S)_C$ as functions $(P/S)_{\text{min}}$, one for each latitude between 20 and 60 degrees latitude. 
Figure~\ref{f:energyapprox} sample monomial approximations at 30 degrees latitude.

\begin{figure}[H]
 \begin{subfigmatrix}{2}% number of columns
     \subfigure[Monomial Approxmiation for $(E/S)_{\text{day}}$\label{f:Benergy}]{\includegraphics{Benergy.pdf}}
     \subfigure[Monomial Approximation for $(E/S)_C$\label{f:Cenergy}]{\includegraphics{Cenergy.pdf}}
 \end{subfigmatrix}
 \caption{\textbf{Monomial approximations for $(E/S)_{\text{day}}$ and $(E/S)_C$ as functions for $(P/S)_{\text{min}}$ at 30 degrees latitude.}}
 \label{f:energyapprox}
\end{figure}

    For the purposes of this design study, it is assumed that the solar cells are can be placed on the wings and the horizontal tail but are not placed on the vertical tail.  
    The fractional solar cell area index $f_{\text{solar}}$ is one when the solar cells completely cover the main wing and greater than one if solar cells are placed on the horizontal tail,

\begin{equation}
    \label{e:solarssolar}
    S_{\text{solar}} \leq f_{\text{solar}}S.
\end{equation}

The total available solar energy per day is an integral of the available solar power during the course of the day,

    \begin{equation}
        \label{e:solares}
        (E/S)_{\text{sun}} = \int_{t_{\text{sun rise}}}^{t_{\text{sun set}}} (P/S)_{\text{sun}} dt.
    \end{equation}

For simplicity, energy and power will refer to energy and power per unit area throughout the rest of this section (i.e. $(P/S) = P$). 
The solar irradiated power $P_{\text{sun}}$ a function of $\theta$, the angle between the normal to the flat surface, or aircraft wing, and the sun beam.\cite{solar}
